Fast gravitational wave parameter estimation without compromises

### 2023-02-07 src/tex/ms.tex
PnP folk added to acks

% Effort going to new tools development.
In order to address this, there are efforts from multiple groups to speed up the
PE process. This includes methods that employ techniques such as
reduced-order quadrature \cite{Canizares:2014fya,Smith:2016qas}, adaptive proposal distributions in nested sampling
\cite{Williams:2021qyt}, deep learning networks pre-trained on large collections
of waveforms \cite{Dax:2021tsq,Dax:2022pxd}, as well as methods that reduce the
computational expense of classical PE by leveraging our knowledge of GW signals
\cite{Veitch:2014wba,Ashton:2021anp,Cornish:2021wxy,Islam:2022afg,Roulet:2022kot,Lee:2022jpn,Lange:2018pyp,Wofford:2022ykb}.
While these avenues are promising for standard GW problems, they rely on
assumptions that may not hold for analyses targeting additional physical effects
beyond standard CBCs in GR, such as lensing and deviations from GR.

@@

In this work, we present a lightweight, flexible, and robust
framework to infer GW event parameters in a fully-Bayesian analysis. Our
framework implements the following techniques to achieve its performance:
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item likelihood heterodyning,
\item differentiable waveform models,
\item a normalizing-flow enhanced Markov chain Monte-Carlo (MCMC) sampler, and
\item native support for hardware accelerators.
\end{enumerate}
All these components, working jointly, come together into a high-performance
and high-fidelity PE pipeline, which can achieve ${\gtrsim}10^3\times$ speed
ups relative to traditional tools without compromising accuracy or generality
for efficiency, and without making limiting assumptions about the target
parameter space.

@@

\section{Acknowledgements}
We thank Will M.~Farr, Aaron Zimmerman, Daniel Foreman-Mackey and Marylou Gabri\'e for helpful discussions.
The Flatiron Institute is a division of the Simons Foundation, supported through the generosity of Marilyn and Jim Simons.
This material is based upon work supported by NSF's LIGO Laboratory which is a major facility fully funded by the National Science Foundation.
This research has made use of data or software obtained from the Gravitational Wave Open Science Center (gw-openscience.org), a service of LIGO Laboratory, the LIGO Scientific Collaboration, the Virgo Collaboration, and KAGRA. LIGO Laboratory and Advanced LIGO are funded by the United States National Science Foundation (NSF) as well as the Science and Technology Facilities Council (STFC) of the United Kingdom, the Max-Planck-Society (MPS), and the State of Niedersachsen/Germany for support of the construction of Advanced LIGO and construction and operation of the GEO600 detector. Additional support for Advanced LIGO was provided by the Australian Research Council. Virgo is funded, through the European Gravitational Observatory (EGO), by the French Centre National de Recherche Scientifique (CNRS), the Italian Istituto Nazionale di Fisica Nucleare (INFN) and the Dutch Nikhef, with contributions by institutions from Belgium, Germany, Greece, Hungary, Ireland, Japan, Monaco, Poland, Portugal, Spain. The construction and operation of KAGRA are funded by Ministry of Education, Culture, Sports, Science and Technology (MEXT), and Japan Society for the Promotion of Science (JSPS), National Research Foundation (NRF) and Ministry of Science and ICT (MSIT) in Korea, Academia Sinica (AS) and the Ministry of Science and Technology (MoST) in Taiwan.

>>>>

% Effort going to new tools development.
In order to address this, there are efforts from multiple groups to speed up the
PE process. This includes methods that employ techniques such as
reduced-order quadrature \cite{Canizares:2014fya,Smith:2016qas}, adaptive proposal distributions in nested sampling
\cite{Williams:2021qyt}, deep learning networks pre-trained on large collections
of waveforms \cite{Dax:2021tsq,Dax:2022pxd}, as well as methods that reduce the
computational expense of classical PE by leveraging our knowledge of GW signals
\cite{Veitch:2014wba,Ashton:2021anp,Cornish:2021wxy,Islam:2022afg,Roulet:2022kot,Lee:2022jpn,Lange:2018pyp,Wofford:2022ykb}.
While these avenues are promising for standard GW problems, they rely on
assumptions that may not hold for analyses targeting additional physical effects
beyond standard CBCs in GR, such as matter effects, lensing and deviations from GR.

@@

In this work, we present a lightweight, flexible, and robust
framework to infer GW event parameters in a fully-Bayesian analysis. Our
approach relies on the following techniques to achieve its performance:
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item likelihood heterodyning,
\item differentiable waveform models,
\item a normalizing-flow enhanced Markov chain Monte-Carlo (MCMC) sampler, and
\item native support for hardware accelerators.
\end{enumerate}
All these components, working jointly, come together into a high-performance
and high-fidelity PE pipeline, which can achieve ${\gtrsim}10^3\times$ speed
ups relative to traditional tools without compromising accuracy or generality
for efficiency, and without making limiting assumptions about the target
parameter space.

@@

\section{Acknowledgements}
We thank Will M.~Farr, Aaron Zimmerman, Daniel Foreman-Mackey and Marylou Gabri\'e for helpful discussions; we also thank Carl-Johan Haster, Neil J.~Cornish and Thomas Dent for comments on the draft.
The Flatiron Institute is a division of the Simons Foundation.
This material is based upon work supported by NSF's LIGO Laboratory which is a major facility fully funded by the National Science Foundation.
This research has made use of data or software obtained from the Gravitational Wave Open Science Center (gw-openscience.org), a service of LIGO Laboratory, the LIGO Scientific Collaboration, the Virgo Collaboration, and KAGRA. LIGO Laboratory and Advanced LIGO are funded by the United States National Science Foundation (NSF) as well as the Science and Technology Facilities Council (STFC) of the United Kingdom, the Max-Planck-Society (MPS), and the State of Niedersachsen/Germany for support of the construction of Advanced LIGO and construction and operation of the GEO600 detector. Additional support for Advanced LIGO was provided by the Australian Research Council. Virgo is funded, through the European Gravitational Observatory (EGO), by the French Centre National de Recherche Scientifique (CNRS), the Italian Istituto Nazionale di Fisica Nucleare (INFN) and the Dutch Nikhef, with contributions by institutions from Belgium, Germany, Greece, Hungary, Ireland, Japan, Monaco, Poland, Portugal, Spain. The construction and operation of KAGRA are funded by Ministry of Education, Culture, Sports, Science and Technology (MEXT), and Japan Society for the Promotion of Science (JSPS), National Research Foundation (NRF) and Ministry of Science and ICT (MSIT) in Korea, Academia Sinica (AS) and the Ministry of Science and Technology (MoST) in Taiwan.
